\startcomponent c_preface
\mullertitle{preface1}{Preface}{to the First Edition.}

The present grammar, which is chiefly intended for beginners, is believed to
contain all the information that a student of Sanskrit is likely to want during
the first two or three years of his reading. Rules referring to the language of
the Vedas have been entirely excluded, for it is not desirable that the
difficulties of that ancient dialect should be approached by any one who has not
fully mastered the grammar of the ordinary Sanskrit such as it was fixed by
Pāṇini and his successors. All allusions to cognate forms in Greek, Latin, or
Gothic, have likewise been suppressed, because, however interesting and useful
to the advanced student, they are apt to deprive the beginner of that clear and
firm grasp of the grammatical system peculiar to the language of ancient India,
which alone can form a solid foundation for the study both of Sanskrit and of
comparative philology.

The two principal objects which I have kept in view while composing this grammar
have been clearness and correctness. With regard to clearness, my chief model
has been the grammar of Bopp; with regard to correctness, the grammar of
Colebrooke. If I may hope, without presumption, to have simplified a few of the
intricacies of Sanskrit grammar which were but partially cleared up by Bopp,
Benfey, Flecchia, and others, I can hardly flatter myself to have reached, with
regard to correctness, the high standard of Colebrooke's great, though
unfinished work. I can only say in self-defence that it is far more difficult to
be correct on every minute point, if one endeavours to rearrange, as I have
done, the materials collected by Pāṇini, and to adapt them to the grammatical
system current in Europe, than if one follows so closely as Colebrooke the
system of native grammarians, and adopts nearly the whole of their technical
terminology. The grammatical system elaborated by native grammarians is, in
itself, most perfect; and those who have tested Pāṇini's work will readily
admit that there is no grammar in any language that could vie with the wonderful
mechanism of his eight books of grammatical rules. But unrivalled as that system
is, it is not suited to the wants of English students, least of all to the wants
of beginners. While availing myself therefore of the materials collected in the
grammar of Pāṇini and in later works, such as the {\em Prakriyā-Kaumudī}, the
{\em Siddhānta-Kaumudī}, the {\em Sārasvatī Prakriyā}, and the {\em
Mādhavīya-dhātu-vṛtti}, I have abstained, as much as possible, from introducing
any more of the peculiar system and of the terminology of Indian
grammarians\footnote{The few alterations that I have made in the usual
terminology have been made solely with a view of facilitating the work of the
learner. Thus instead of numbering the ten classes of verbs, I have called each
by its first verb. This relieves the memory of much unnecessary trouble, as the
very name indicates the character of each class; and though the names may at
first sound somewhat uncouth, they are after all the only names recognized by
native grammarians. Knowing from my experience as an examiner how difficult it
is to remember the merely numerical distinction between the first, second, or
third preterites, or the first and second futures, I have kept as much as
possible to the terminology with which classical scholars are familiar, calling
the tense corresponding to the Greek imperfect, {\em imperfect}; that
corresponding to the perfect, {\em reduplicated perfect}; that corresponding to
the aorist, {\em aorist}; and the mood corresponding to the optative, {\em
optative}. The names of {\em periphrastic perfect} and {\em periphrastic future}
tell their own story; and if I have retained the merely numerical distinction
between the first and second aorists, it was because this distinction seemed to
be more intelligible to a classical scholar than the six or seven forms of the
so-called {\em multiform preterite}. If it were possible to make a change in the
established grammar nomenclature, I should much prefer to call the {\em first}
the {\em second}, and the {\em second} the {\em first aorist}; the former being
a secondary and compound, the latter a primary and simple tense. But {\em first}
and {\em second aorists} have become almost proper names, and will not easily
yield their place to mere argument.} than has already found admittance into our
Sanskrit grammars; nay, I have frequently rejected the grammatical observations
supplied ready to hand in their works in order not to overwhelm the memory of
the student with too many rules and too many exceptions. Whether I have always
been successful in drawing a line between what is essential in Sanskrit grammar
and what is not, I must leave to the judgement of those who enjoy the good
fortune of being engaged in the practical teaching of a language the students of
which may be counted no longer by tens, but by hundreds.\footnote{In the
University of Leipsig alone, as many as fifty pupils attended every year the
classes of Professor Brockhaus in order to acquire a knowledge of the elements
of Sanskrit, previous to the study of comparative philology under Professor
Curtius.} I only wish it to be understood that where I have left out rules or
exceptions, contained in other grammars, whether native or European, I have done
so after mature consideration, deliberately preferring the less complete to the
more complete, but, at the same time, more bewildering statement of the
anomalies of the Sanskrit language. Thus, to mention one or two cases, when
giving the rules on the employment of the suffixes \tl{vat} and \tl{mat} (§187),
I have left out the rule that bases ending in \tl{m}, though the \tl{m} be
preceded by other vowels than \tl{a}, always take \tl{vat} instead of \tl{mat}.
I did so partly because there are very few bases ending in {\tl m}, partly
because, if a word like \tl{kim-vān} should occur, it would be easy to discover
the reason why here too \tl{v} was preferred to \tl{m}, viz.\ %
% SECOND EDITION:
in order to avoid the clashing of two \tl{m}'s.
% FIRST EDITION:
% that bases ending in {\tl m} are not allowed to form denominatives.
%
% SECOND EDITION:
I passed over, for very much the same reason, the prohibition given in Pāṇini
3.8.3, viz.\ that bases ending in \tl{m} are not allowed to form
denominatives.
% end of addition
It is true, no doubt, that the omission of such rules or exceptions may be said
to involve an actual misrepresentation, and that a pupil might be mislead to
form such words as \tl{kim-mān} and \tl{kim-yati}. But this cannot be avoided in
an elementary grammar; and the student who is likely to come into contact with
such recondite forms will no doubt be sufficiently advanced to be able to
consult for himself the rules of Pāṇini and the explanations of his
commentators.

My own fear is that, in writing an elementary grammar, I have erred rather in
giving too much than in giving too little. I have therefore in the table of
contents marked with an asterisk all such rules as may be safely left out in a
first course of Sanskrit grammar,%
% second edition
\footnote{In the second edition all these paragraphs are printed in smaller
type.}
 % keep this space
and I have in different places informed the reader whether certain portions
might be passed over quickly, or should be carefully committed to memory. Here
and there, as for instance in §103, a few extracts are introduced from Pāṇini
simply in order to give to the student a foretaste of what he may expect in the
elaborate works of native grammarians, while lists of verbs like those contained
in §332 or §462 are given, as everybody will see, for the sake of reference
only. The somewhat elaborate treatment of the nominal bases in \tl{ī} and
\tl{ū}, from §220 to §226, became necessary partly because in no grammar had the
different paradigms of this class been correctly given, partly because it was
impossible to bring out clearly the principle on which the peculiarities and
apparent irregularities of these nouns are based without entering fully into the
systematic arrangement of native grammarians. Of portions like this I will not
say indeed, {μωμήσεταί τις μᾶλλον ἢ μιμήσεται}, but I feel that I may say,
{\ss यत्ने कृते यदि न सिघ्यति कीऽत्न दोषः}; and I know that those who will take the trouble
to examine the same mass of evidence which I have weighed and examined will be
the most lenient in their judgement, if hereafter they should succeed better
than I have done in unravelling the intricate argumentations of native scholars.
\footnote{To those who have the same faith in the accurate and never swerving
argumentations of Sanskrit commentators, it may be a saving of time to be
informed that in the new and very useful edition of the {\em Siddhānta-Kaumudī}
by Śrī Tārānātha-tarkavā-caspati there are two misprints which hopelessly
disturb the order of the rules on the proper declension of nouns in \tl{ī} and
\tl{ū}. On page~136, line~7, read {\ss श्रीवत्‌} instead of {\ss स्त्रीवत्‌}; this is
corrected in the {\em Corrigenda}, and the right reading is found in the old
edition. On the same page, line~13, insert {\ss न} after {\ss विना}, or join
{\ss विनास्त्रीबोघकत्वं}.}

But while acknowledging my obligations to the great grammarians of India, it
would be ungrateful were I not to acknowledge as fully the assistance which I
have derived from the works of European scholars. My first acquaintance with the
elements of Sanskrit was gained from Bopp's grammar. Those only who know the
works of his predecessors, of Colebrooke, Carey, Wilkins, and Forster, can
appreciate the advance made by Bopp in explaining the difficulties, and in
lighting up, if I may say so, the dark lanes and alleys of the Sanskrit
language. I doubt whether Sanskrit scholarship would have flourished as it has,
if students had been obliged to learn their grammar from Forster or Colebrooke,
and I believe that to Bopp's little grammar is due a great portion of that
success which has attended the study of Sanskrit literature in Germany.
Colebrooke, Carey, Wilkins, and Forster worked independently of each other. Each
derived his information from native teachers and from native grammars. Among
these four scholars, Wilkins seems to have been the first to compose a Sanskrit
grammar, for he informs us that the first printed sheet of his work was
destroyed by fire in 1795. The whole grammar, however, was not published till
1808. In the mean time Forster had finished his grammar, and had actually
delivered his MS to the Council of the College of Fort William in 1804. But it
was not published till 1810. The first part of Colebrooke's grammar was
published in 1805, and therefore stands first in point of time of publication.
Unfortunately it was not finished, because the grammars of Forster and Carey
were then in course of publication, and would, as Colebrooke imagined, supply
the deficient part of his own. Carey's grammar was published in 1806. Among
these four publications, which as first attempts at making the ancient language
of India accessible to European scholars, deserve the highest credit,
Colebrooke's grammar is {\em facile princeps}. It is derived at first hand from
the best native grammars, and evinces a familiarity with the most intricate
problems of Hindu grammarians such as few scholars have acquired after him. No
one can understand and appreciate the merits of this grammar who has not
previously acquired a knowledge of the grammatical system of Pāṇini, and it is a
great loss to Sanskrit scholarship that so valuable a work should have remained
unfinished.

I owe most, indeed, to Colebrooke and Bopp, but I have derived many useful hints
from other grammars also. There are some portions of Wilson's grammar which show
that he consulted native grammarians, and the fact that he possessed the
remaining portion of Colebrooke's\footnote{See Wilson's {\em Sanscrit and
English Dictionary}, first edition, preface, p.\ xlv.} MS gives to his list of
verbs, with the exception of the \tl{bhū} class, which was published by
Colebrooke, a peculiar interest. Professor Benfey in his large grammar performed
a most useful task in working up independently the materials supplied by Pāṇini
and Bhaṭṭoji Dīkṣita; and his smaller grammars too, published both in German and
in English, have rendered good service to the cause of sound scholarship. There
are besides, the grammars of Boller in German, of Oppert in French, of
Westergaard in Danish, of Flecchia in Italian, each supplying something that
could not be found elsewhere, and containing suggestions, many of which have
proved useful to the writer of the present grammar.

But while thus rendering full justice to the honest labours of my predecessors,
I am bound to say, at the same time, that with regard to doubtful or difficult
forms, of which there are many in the grammar of the Sanskrit language, not one
of them can be appealed to as an ultimate authority. Every grammar contains, as
is well known, a number of forms which occur but rarely, if ever, in the
literary language. It is necessary, however, for the sake of systematic
completeness to give these forms; and if they are to be given at all, they must
be given on competent authority. Now it might be supposed that a mere reference
to any of the numerous grammars already published would be sufficient for this
purpose, and that the lists of irregular or unusual forms might safely be copied
from their pages. But this is by no means the case. Even with regard to regular
forms, whoever should trust implicitly in the correctness of any of the grammars
hitherto published would never be certain of having the right form. I do not say
this lightly or without being able to produce proofs. When I began to revise my
manuscript grammar which I had composed for my own use many years ago, and when
on points on which I felt doubtful, I consulted other grammars, I soon
discovered either that, with a strange kind of sequacity, they all repeated the
same mistake, or that they varied wildly from each other without assigning any
reason or authority. I need not say that the grammars which we possess differ
very much in the degree of their trustworthiness; but with the exception of the
first volume of Colebrooke and of Professor Benfey's larger Sanskrit grammar, it
would be impossible to appeal to any of my predecessors as an authority on
doubtful points. Forster and Carey, who evidently depend almost entirely on
materials supplied to them by native assistants, give frequently the most
difficult forms with perfect accuracy, while they go wildly wrong immediately
after, without, it would seem, any power of controlling their authorities. The
frequent inaccuracies in the grammars of Wilkins and Wilson have been pointed
out by others; and however useful these works may have been for practical
purposes, they were never intended as authorities on contested points of
Sanskrit grammar.

Nothing remained, in fact, in order to arrive at any satisfactory result, but to
collate the whole of my grammar with regard not only to the irregular but
likewise to the regular forms, with Pāṇini and other native grammarians, and to
supply for each doubtful case, and for rules that might seem to differ from
those of any of my predecessors, a reference to Pāṇini or to other native
authorities. This I have done, and in so doing I had to rewrite nearly the whole
of my grammar; but though the time and trouble expended on this work have been
considerable, I believe that they have not been bestowed in vain. I only regret
that I did not give these authoritative references throughout the whole of my
work,%
% second edition
\footnote{They have been in the second edition.}
 %
because, even where there cannot be any difference of opinion, some of my
readers might thus have been saved the time and trouble of looking through
Pāṇini to find the \tl{sūtras} that bear on every form of the Sanskrit
language.

By this process which I have adopted, I believe that on many points a more
settled and authoritative character has been imparted to the grammar of Sanskrit
than it possessed before; but I do by no means pretend to have arrived on all
points at a clear and definite view of the meaning of Pāṇini and his
successors. The grammatical system of Hindu grammarians is so peculiar that
rules which we should group together are scattered about in different parts of
their manuals. We may have the general rule in the last, and the exceptions in
the first book, and even then we are by no means certain that exceptions to
these exceptions may not occur somewhere else. I shall give but one instance.
There is a root \tr{जागृ}{jāgṛ}, which forms its aorist by adding
\tr{इषम्}{iṣam}, \tr{ईः}{īḥ}, \tr{ईत्}{īt}. Here the simplest rule would be that
final \tr{ऋ}{ṛ} before \tr{इषम्}{iṣam} becomes \tr{र्}{r} \panini{6.1.77}. This,
however, is prevented by another rule which requires that final \tr{ऋ}{ṛ} should
take guṇa before  \tr{इषम्}{iṣam} \panini{7.3.84}. This would give us
\tr{अजागरिषम्}{ajāgar-iṣam}. But now comes another general rule \panini{7.2.1}
which prescribes vṛddhi of final vowels before \tr{इषम्}{iṣam}, i.e.\
\tr{अजागारिषम्}{ajāgāriṣam}. Against this change, however, a new rule is cited
\panini{7.3.85}, and this secures for \tr{जागृ}{jāgṛ} a special exception from
vṛddhi, and leaves its base again as  \tr{जागर्}{jāgar}. As soon as the base has
been changed to  \tr{जागर्}{jāgar}, it falls under a new rule \panini{7.2.3},
and is forced to take vṛddhi until this rule is again nullified by Pāṇini
7.2.4, which does not allow vṛddhi in an aorist that takes intermediate
\tr{इ}{i}, like \tr{अजागरिषम्}{ajāgariṣam}. There is an exception, however, to
this rule also for bases with short \tr{अ}{a} beginning and ending with a
consonant may optionally take vṛddhi \panini{7.2.7}. This option is afterwards
restricted, and roots with short \tr{अ}{a}, beginning with a consonant and
ending in \tr{र्}{r}, like \tr{जागर्}{jāgar}, have no option left, but are
restricted afresh to vṛddhi \panini{7.2.2}. However, even this is not yet the
final result. Our base  \tr{जागर्}{jāgar} is after all not to take vṛddhi, and
hence a new special rule \panini{7.2.5} settles the point by granting to
\tr{जागृ}{jāgṛ} a special exception from vṛddhi and thereby establishing its
guṇa. No wonder that these manifold changes and chances in the formation of the
first aorist of \tr{जागृ}{jāgṛ} should have inspired a grammarian, who
celebrates them in the following couplet:

\setupnarrower[left=1.5in]
\blank\startnarrower
  {\ss गुणो वृद्धिर्गुणो वृद्धिः प्रतिषेधो विकल्पनं ।

  पुनर्वृद्धिर्निषेधोऽतो यण्‌पूर्वाः प्राप्रयो नव ।।}
\stopnarrower\blank

\quotation{Guṇa, vṛddhi, guṇa, vṛddhi, prohibition, option, again vṛddhi and
then exception, these, with the change of \tl{ṛ} into a semivowel in the first
instance, are the nine results.}

Another difficulty consists in the want of critical accuracy in the editions
which we possess of Pāṇini, the {\em Siddhānta-Kaumudī}, the {\em
Laghu-Kaumudī}, the {\em Sārasvatī}, and Vopadeva. Far be it from me to wish to
detract from the merits of native editors, like Dharaṇīdhara, Kāśīnātha,
Tārānātha, still less from those of Professor Boehtlingk, who published his text
and notes nearly thirty years ago, when few of us were able to read a single
line of Pāṇini. But during those thirty years considerable progress has been
made in unravelling the mysteries of the grammatical literature of India. The
commentary of Sāyaṇa to the Rigveda has shown us how practically to apply the
rules of Pāṇini; and the translation of the {\em Laghu-Kaumudī} by the late Dr
Ballantyne has enabled even beginners to find their way through the labyrinth of
native grammar. The time has come, I believe, for new and critical editions of
Pāṇini and his commentators. A few instances may suffice to show the insecurity
of our ordinary editions. The commentary to Pāṇini 7.2.42, as well as the {\em
Sārasvatī} 2.25.1, gives the benedictive ātmanepada \tr{वरीषीष्ट}{varīṣīṣṭa} and
\tr{स्तरीषीष्ट}{starīṣīṣṭa}; yet a reference to Pāṇini 7.2.39 and 40 shows that
these forms are impossible. Again, if Pāṇini (8.3.92) is right%
% second edition
---and how could the Infallible be wrong?---%
%
in using \tr{अग्रगामिनि}{agragāmini} with a dental \tl{n} in the last
syllable, it is clear that he extends the prohibition given in 8.4.34, with
regard to \tl{upasargas}, to other compounds. It is useless to inquire whether
in doing so he was right or wrong, for it is an article of faith with every
Hindu grammarian that whatever word is used by Pāṇini in his \tl{sūtras} is
{\em eo ispo} correct. Otherwise, the rules affecting compounds with
\tl{upasargas} are by no means identical with those that affect ordinary
compounds; and though it may be right to argue {\em a fortiori} from
\tr{प्रगामिनि}{pragāmini} to \tr{अग्रगामिनि}{agragāmini}, it would not be right
to argue from \tr{अग्रयान}{agrayāna} to \tr{प्रयान}{prayāna}, this being
necessarily \tr{प्रयाण}{prayāṇa}. But assuming  \tr{अग्रगामिनि}{agragāmini} to
be correct, it is quite clear that the compounds
\tr{स्वर्गकामिणौ}{svargakāmiṇau},  \tr{वृषगामिणौ}{vṛṣagāmiṇau},
\tr{हरिकामाणि}{harikāmāṇi}, and  \tr{हरिकामेण}{harikāmeṇa}, given in the
commentary to 8.4.13, are all wrong, though most of them occur not only in the
printed editions of Pāṇini and the {\em Siddhānta-Kaumudī}, but may be traced
back to the MSS of the {\em Prakriyā-Kaumudī}, the source, though by no means
the model, of the {\em Siddhānta-Kaumudī}. I was glad to learn from my friend
Professor Goldstücker, who is preparing an edition of the {\em Kāśikāvṛtti}, and
whom I consulted on these forms, that the MSS of Vāmana which he possesses
carefully avoid these faulty examples to Pāṇini 8.4.13.

After these explanations I need hardly add that I am not so sanguine as to
suppose that I could have escaped scot free where so many men of superior
knowledge and talent have failed to do so. All I can say is that I shall be
truly thankful to any scholar who will take the trouble to point out any
mistakes into which I may have fallen; and I hope that I shall never so far
forget the regard due to truth as to attempt to represent simple corrections,
touching the declension of nouns or the conjugation of verbs, as matters of
opinion, or so far lower the character of true scholarship as to appeal, % keep
space
% second edition
on such matters, %
%
from the verdict of the few to the opinion of the many.

Hearing from my friend Professor Bühler that he had finished a Sanskrit syntax,
based on the works of Pāṇini and other native grammarians, which will soon be
published, I gladly omitted that portion of my grammar. The rules on the
derivation of nouns by means of \tl{kṛt}, \tl{uṇādi}, and \tl{taddhita}
suffixes do not properly belong to the sphere of an elementary grammar. If time
and health permit, I hope to publish hereafter, as a separate treatise, the
chapter of the {\em Prakriyā-Kaumudī} bearing on this subject.

In the list of verbs which I have given as an appendix, pp.\ 245--299, I have
chiefly followed the {\em Prakriyā-Kaumudī} and the {\em Sārasvatī}. These
grammars do not conjugate every verb that occurs in the {\em Dhātupāṭha}, but
those only that serve to illustrate certain grammatical rules. Nor do they
adopt, like the {\em Siddhānta-Kaumudī}, the order of the verbs as given in
Pāṇini's {\em Dhātupāṭha}, but they group the verbs of each class according to
their voices, treating together those that take the terminations of the
parasmaipada, those that take the terminations of the ātmanepada, and, lastly,
those that admit of both voices. In each of these subdivisions, again, the
simple verbs are so arranged as best to illustrate certain grammatical rules. In
making a new selection among the verbs selected by Rāmachandra and
Anubhūtisvarūpāchārya, I have given a preference to those which occur more
frequently in Sanskrit literature and to those which illustrate some points of
grammar of peculiar interest to the student. In this manner I hope that the
appendix will serve two purposes: it will not only help the student, when
doubtful as to the exact forms of certain verbs, but it will likewise serve as a
useful practical exercise to those who, taking each verb in turn, will try to
account for the exact forms of its persons, moods, and tenses by a reference to
the rules of this grammar. In some cases references have been added to guide the
student, in others he has to find by himself the proper warranty for each
particular form.

My kind friends Professor Cowell and Professor Kielhorn have revised some of the
proof-sheets of my grammar, for which I beg to express to them my sincere
thanks.

\blank
\startnarrower
\startalignment[flushright]
% second edition
F. %
%
MAX MÜLLER.
\stopalignment
\blank

{\tfx PARIS,

5th April, 1866.}
\stopnarrower
\stopcomponent
