\startcomponent c_preface2ed
\mullertitle{preface2}{Preface}{to the Second Edition.}

The principal alterations in the new edition of my Sanskrit grammar consist in a
number of additional references to Pāṇini, in all cases where an appeal to his
authority seemed likely to be useful, and in the introduction of the marks of
the accent. I have also been able to remove a number of mistakes and misprints
which, in spite of all the care I had taken, had been overlooked in the first
edition. Most of these I had corrected in the German translation of my grammar,
published at Leipzig in 1868; some more have now been corrected. I feel most
grateful to several of my reviewers for having pointed out these oversights, and
most of all to Pandit Rājārāmaśāstrī, whose list of notes and queries to my
grammar has been of the greatest value to me. It seems almost hopeless for a
European scholar to acquire that familiarity with the intricate system of Pāṇini
which the Pandits of the old school in India still possess; and although some of
their refinements in the interpretation of Pāṇini's rules may seem too subtle,
yet there can be no doubt that these living guides are invaluable to us in
exploring the gigantic labyrinth of ancient Sanskrit grammar.

There is, however, one difficulty which we have to contend with, and which does
not exist for them. They keep true throughout to one system, the system of
Pāṇini; we have to transfer the facts of that system into our own system of
grammar. What accidents are likely to happen during this process I shall try to
illustrate by one instance. Rājārāmaśāstrī objects to the form \tr{पुन्सु}{punsu}
as the locative plural of \tr{पुमान}{pumān}. From his point of view, he is
perfectly right in his objection, for according to Pāṇini the locative plural
has anusvāra, \tr{पुंसु}{puṁsu}. But in our own Sanskrit grammars we first have a
general rule that \tr{स्}{s} is changed to \tr{ष्}{ṣ} after any vowel except
\tr{अ}{a}, in spite of intervening anusvāra (see §100); and it has even been
maintained that there is some kind of physiological reason for such a change. If
then, after having laid down this rule, we yet write \tr{पुंसु}{puṁsu}, we simply
commit a grammatical blunder; and I believe there is no Sanskrit grammar, except
Colebrooke's, in which that blunder has not been committed. In order to avoid
it, I wrote \tr{पुन्सु}{punsu}, thus, by the retention of the dental \tr{न}{n},
making it grammatically and physically possible for the \tr{स्}{s} to remain
unchanged. It may be objected that on the same ground I ought to have written
instrumental \tr{पुन्सा}{punsā}, genitive \tr{पुन्सः}{punsaḥ}, \&c.; but in these
cases the \tr{स्}{s} is radical, and would therefore not be liable to be changed
into \tr{ष्}{ṣ} after a vowel and anusvāra \panini{8.3.59}. Professor Weber had
evidently overlooked these simple rules, or he would have been less forward in
blaming Dr Keller for having followed my example in writing \tr{पुन्सु}{punsu},
instead of \tr{पुंसु}{puṁsu}. In Pāṇini's grammar (as may be seen from my note
appended to §100) the rule on the change of \tr{स्}{s} into \tr{ष्}{ṣ} is so
carefully worded that it just excludes the case of \tr{पुंसु}{puṁsu}, although the
\tr{सु}{su} of the locative plural is preceded by an anusvāra. I have now, by
making in my second edition the same reservation in the general rule, been able
to conform to Pāṇini's authority, and have written \tr{पुंसु}{puṁsu}, instead of
\tr{पुन्सु}{punsu}, though even thus the fact remains that if the dot is really
meant for anusvāra, and if the \tr{सु}{su} is the termination of the locative
plural, the \tr{स्}{s} would be sounded as \tr{ष्}{ṣ}, according to the general
tendency of the ancient Sanskrit pronunciation.

I have mentioned this one instance in order to show the peculiar difficulties
which the writer of a Sanskrit grammar has to contend with in trying to combine
the technical rules of Pāṇini with the more rational principles of European
grammar; and I hope it may convince my readers, and perhaps even Professor
Weber, that where I have deviated from the ordinary rules of our European
grammars, or where I seem to have placed myself at variance with some of the
native authorities, I have not done so without having carefully weighed the
advantages of the one against those of the other system.

\blank
\startnarrower
\rightaligned{F. MAX MÜLLER.}

\blank
{\tfx PARKS END, OXFORD,

August, 1870.}
\stopnarrower
\stopcomponent
